---
title: "cori.data.fcc"
description: "R package for accessing and analyzing FCC broadband data"
author: "CORI MDA Team"
date: "2025-11-20"
categories: ["R Package", "Broadband"]
tags: ["FCC", "Data Processing", "R", "National Broadband Map"]
package:
  name: "cori.data.fcc"
  githubUrl: "https://github.com/ruralinnovation/cori.data.fcc"
  installCommand: 'remotes::install_github("ruralinnovation/cori.data.fcc")'
  status: "stable"
  version: "0.2.1"
  maintainer: "CORI MDA Team"
  featured: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Overview

The `cori.data.fcc` package provides tools for downloading, processing, and analyzing FCC broadband data in R. It simplifies access to the National Broadband Map and other FCC data sources, making it easier to conduct broadband equity and availability analysis.

::: callout-note
## Why Use This Package

- **Simplified data access** - Download FCC data directly from R
- **Standardized processing** - Consistent data cleaning and formatting
- **Analysis tools** - Built-in functions for coverage gaps and equity analysis
- **Integration ready** - Works seamlessly with CORI data pipelines
:::

## Installation

::: panel-tabset
## GitHub

```{r}
#| eval: false

# Install remotes if you haven't already
install.packages("remotes")

# Install cori.data.fcc
remotes::install_github("ruralinnovation/cori.data.fcc")
```

## Dependencies

The package requires standard R data manipulation libraries:

```{r}
#| eval: false

install.packages(c("dplyr", "tidyr", "sf", "httr"))
```
:::

## Quick Start

```{r}
#| eval: false

library(cori.data.fcc)

# Download broadband availability data for a state
data <- get_broadband_data(state = "VT")

# Analyze coverage by technology type
coverage_summary <- summarize_coverage(data, by = "technology")

# Create coverage map
plot_coverage(data, metric = "download_speed")
```

## Key Functions

### Data Access

```{r}
#| eval: false

# Download broadband availability data
get_broadband_data(
  state = "VT",           # Two-letter state code
  geography = "tract",    # "tract", "block", or "county"
  vintage = "latest"      # FCC data version
)
```

### Analysis Functions

```{r}
#| eval: false

# Aggregate coverage statistics
summarize_coverage(
  data,
  by = "technology"      # "technology", "provider", or "speed_tier"
)

# Identify coverage gaps
calculate_gaps(
  data,
  threshold = 25/3       # Download/upload speed threshold in Mbps
)
```

### Visualization

```{r}
#| eval: false

# Create coverage visualizations
plot_coverage(
  data,
  metric = "download_speed",    # Metric to visualize
  geography = "tract"            # Geographic level
)
```

## Common Use Cases

### State-Level Coverage Analysis

```{r}
#| eval: false

library(dplyr)

# Get Vermont broadband data
vt_data <- get_broadband_data(state = "VT")

# Calculate coverage rates
coverage_rates <- vt_data |>
  summarize_coverage(by = "technology") |>
  arrange(desc(coverage_pct))
```

### Identifying Coverage Gaps

```{r}
#| eval: false

# Find areas lacking 25/3 Mbps service
gaps_25_3 <- calculate_gaps(
  vt_data,
  threshold = list(download = 25, upload = 3)
)

# Find areas lacking 100/20 Mbps service
gaps_100_20 <- calculate_gaps(
  vt_data,
  threshold = list(download = 100, upload = 20)
)
```

### Technology Comparison

```{r}
#| eval: false

# Compare fiber vs. cable availability
tech_comparison <- vt_data |>
  filter(technology %in% c("Fiber", "Cable")) |>
  group_by(technology) |>
  summarize(
    avg_download = mean(download_speed, na.rm = TRUE),
    avg_upload = mean(upload_speed, na.rm = TRUE),
    coverage_pct = sum(covered) / n() * 100
  )
```

## Data Structure

**Standard Output Format:**

- One row per geographic unit per provider per technology
- Standardized variable names
- Geographic identifiers (GEOID)
- Speed tiers and technology types

**Typical Variables:**

```
geoid, state, county, tract, block, provider_name, technology,
download_speed, upload_speed, covered, available_speeds
```

## Integration with CORI Workflows

### With Rural Definitions

```{r}
#| eval: false

library(ruraldefinitions)

# Get broadband data
broadband <- get_broadband_data(state = "VT")

# Add rural classification
broadband_rural <- broadband |>
  left_join(
    get_definition("ruca", 2020),
    by = "geoid"
  )

# Compare rural vs. urban coverage
coverage_by_rural <- broadband_rural |>
  group_by(is_rural) |>
  summarize_coverage()
```

### With Census Data

```{r}
#| eval: false

library(tidycensus)

# Get population data
pop_data <- get_acs(
  geography = "tract",
  variables = "B01001_001",
  state = "VT"
)

# Join with broadband data
broadband_pop <- broadband |>
  left_join(pop_data, by = "geoid") |>
  mutate(
    unserved_population = population * (1 - coverage_pct / 100)
  )
```

## Use Cases at CORI

This package is used in:

- [Broadband Equity Analysis Project](/projects/broadband-equity/)
- [FCC Data Overview Blog Post](/stories/fcc-data-overview/)
- [Estimating Broadband Adoption Impact](/stories/estimating-adoption-impact/)
- Rural broadband gap identification
- BEAD (Broadband Equity, Access, and Deployment) planning

## Data Sources

The package accesses:

- **FCC National Broadband Map** - Broadband availability and speed data
- **FCC Form 477** - Provider coverage submissions (historical)
- **BDC (Broadband Data Collection)** - Updated availability data (current)

## Performance Considerations

::: callout-tip
## Optimization Tips

1. **Download by state** - Reduces data transfer and processing time
2. **Use appropriate geography** - County is faster than block for large-scale analysis
3. **Filter early** - Subset data before complex calculations
4. **Cache results** - Save processed data locally for repeated analysis
:::

## Known Limitations

::: callout-warning
## Important Notes

- **Coverage vs. Adoption** - Data shows availability, not actual subscriptions
- **Provider reporting** - Based on provider submissions, may contain inaccuracies
- **Update lag** - Data may be 6-12 months behind current conditions
- **Block-level challenges** - Census blocks can be very large in rural areas
:::

## Related Resources

### Datasets

- [FCC National Broadband Map](/datasets/fcc-broadband/) - Source dataset documentation

### Related Packages

- [ruraldefinitions](/packages/ruraldefinitions/) - Rural classification definitions
- `tidycensus` - Census demographic data
- `sf` - Spatial data manipulation

### External Links

- [FCC National Broadband Map](https://broadbandmap.fcc.gov/)
- [FCC BDC Data Downloads](https://www.fcc.gov/BroadbandData)
- [NTIA Broadband Resources](https://broadbandusa.ntia.doc.gov/)

## Documentation

- [GitHub Repository](https://github.com/ruralinnovation/cori.data.fcc)
- [Function Reference](https://github.com/ruralinnovation/cori.data.fcc/wiki)
- [Example Vignettes](https://github.com/ruralinnovation/cori.data.fcc/tree/main/vignettes)

---

**Contact:** CORI MDA Team

For bug reports or feature requests, please open an issue on [GitHub](https://github.com/ruralinnovation/cori.data.fcc/issues).
