---
title: "FCC National Broadband Map"
description: "Comprehensive broadband availability and adoption data from the Federal Communications Commission"
author: "CORI Data Team"
date: "2025-12-15"
categories: ["Dataset", "Broadband", "Infrastructure"]
tags: ["FCC", "Connectivity", "Geographic"]
dataset:
  name: "FCC National Broadband Map"
  source: "Federal Communications Commission"
  sourceUrl: "https://broadbandmap.fcc.gov/data-download"
  accessMethod: "Direct Download and API"
  updateFrequency: "Biannual"
  geographicLevel: "Census Block, County, State"
  dataFormat: ["CSV", "Parquet"]
  featured: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Overview

The FCC National Broadband Map provides detailed information about broadband availability across the United States. This dataset includes information on:

- Broadband service provider coverage
- Technology types (Fiber, Cable, DSL, etc.)
- Upload and download speeds
- Census block-level granularity

The National Broadband Map is the authoritative source for understanding broadband access gaps and measuring progress toward universal connectivity.

::: callout-note
## Why FCC Broadband Data is Essential

- **Location-level precision** - Census block level data (finest geographic detail)
- **Technology detail** - Fiber, cable, DSL, fixed wireless, satellite
- **Speed tiers** - Upload and download speeds by provider
- **Biannual updates** - More frequent than survey-based measures
:::

## Coverage

::: columns
::: {.column width="50%"}
**Geographic Coverage**

- Census block level
- County aggregations
- State aggregations
- All 50 states
:::

::: {.column width="50%"}
**Temporal Coverage**

- Updated twice per year
- Historical data available
- Provider-reported service availability
- Challenge process for data verification
:::
:::

## Key Features

### Geographic Granularity

Census block-level data allows for:

- Precise gap analysis
- Rural vs. urban comparisons
- Infrastructure planning
- Grant targeting (BEAD, ReConnect, etc.)

### Technology Types

::: panel-tabset
## Wireline Technologies

- **Fiber** - Fiber to the premises (FTTP/FTTH)
- **Cable** - Hybrid fiber-coaxial (HFC)
- **DSL** - Digital subscriber line
- **Copper** - Legacy telephone lines

## Fixed Wireless

- **Licensed Fixed Wireless** - Point-to-multipoint systems
- **Unlicensed Fixed Wireless** - WiFi-based systems
- **5G Home Internet** - Cellular-based home broadband

## Satellite

- **Geostationary Satellite** - Traditional satellite (high latency)
- **Low Earth Orbit (LEO) Satellite** - Starlink, OneWeb, etc.
:::

### Speed Tiers

Upload and download speeds reported by providers:

- 25/3 Mbps (legacy broadband definition)
- 100/20 Mbps (BEAD threshold)
- 1 Gbps+ (gigabit service)

## Usage

This dataset is commonly used in:

::: panel-tabset
## Broadband Equity

- Identifying unserved and underserved locations
- BEAD (Broadband Equity, Access, and Deployment) program
- Measuring digital divide

## Rural Connectivity

- Rural broadband availability
- Technology adoption patterns
- Infrastructure investment targeting

## Infrastructure Planning

- Network deployment planning
- Cost modeling for buildout
- Service area mapping

## Policy Research

- Measuring progress toward universal broadband
- Evaluating broadband programs
- Competition analysis
:::

## Access

### FCC Broadband Map Download

::: panel-tabset
## Direct Download

- [FCC Broadband Map Download](https://broadbandmap.fcc.gov/data-download)

## CORI S3 Bucket

- [CORI S3 Bucket](https://s3.amazonaws.com/cori-data/fcc/)

## cori.data.fcc R Package

```{r}
#| eval: false

# Using cori.data.fcc package
library(cori.data.fcc)

# Get broadband availability data
fcc_data <- get_fcc_broadband(
  geography = "county",
  vintage = "2024-06"
)
```
:::

## Data Structure

**Census Block Format:**

- One row per census block per provider per technology
- Detailed service characteristics
- Maximum advertised speeds

**County Aggregated Format:**

- Unserved/underserved location counts
- Technology availability percentages
- Provider competition metrics

## Methodology

### Provider Reporting

Broadband service providers report coverage data to the FCC:

- Service availability by census block
- Maximum advertised speeds
- Technology types deployed
- Reported twice per year

### Challenge Process

::: callout-important
## Data Verification

The FCC uses a challenge process to verify provider-reported data:

- State and local governments can challenge data
- Service providers must respond to challenges
- Data updated based on verified challenges
- Improves accuracy over time
:::

### Fabric Data

The FCC uses "Fabric" data:

- Comprehensive list of all broadband-serviceable locations (BSLs)
- More accurate than census block aggregation
- Enables location-level analysis

## Strengths & Limitations

::: columns
::: {.column width="50%"}
### Strengths

- Census block precision
- Technology detail
- Speed tier information
- Challenge process improves accuracy
- Biannual updates
:::

::: {.column width="50%"}
### Limitations

- Provider self-reporting (verification ongoing)
- Availability ≠ adoption
- No pricing information
- No quality/reliability metrics
- Lag in challenge resolution
:::
:::

## Important Notes

### Availability vs. Adoption

::: callout-warning
## Critical Distinction

**Availability:** Broadband service is available at a location (from FCC map)

**Adoption:** Households subscribe to broadband service (from ACS)

FCC data measures **availability**, not actual subscriptions. Use [ACS](/datasets/american-community-survey/) for adoption rates.
:::

### Speed Definitions

Different programs use different speed thresholds:

| Program | Download | Upload | Definition |
|---------|----------|--------|------------|
| Legacy FCC | 25 Mbps | 3 Mbps | "Broadband" |
| BEAD | 100 Mbps | 20 Mbps | "Underserved" |
| BEAD | <25 Mbps | <3 Mbps | "Unserved" |

### Technology Considerations

Not all technologies are equal:

- **Fiber:** Future-proof, symmetrical speeds, low latency
- **Cable:** High download, lower upload, shared bandwidth
- **DSL:** Distance-limited, slower speeds
- **Fixed Wireless:** Weather-dependent, line-of-sight required
- **Satellite:** High latency (except LEO), expensive

## Related Resources

- [cori.data.fcc R Package](/packages/cori-data-fcc/)
- [FCC Data Overview Blog Post](/stories/fcc-data-overview/)
- [Broadband Equity Project](/projects/broadband-equity/)

## R Packages

### Recommended Packages

```{r}
#| eval: false

# Install cori.data.fcc package
# remotes::install_github("ruralinnovation/cori.data.fcc")
library(cori.data.fcc)
```

### Usage Example

```{r}
#| eval: false

library(cori.data.fcc)

# Get county-level broadband availability
county_broadband <- get_fcc_broadband(
  geography = "county",
  vintage = "2024-06",
  speeds = c("100_20", "25_3")
)
```

## Additional Resources

- [FCC National Broadband Map](https://broadbandmap.fcc.gov/)
- [BEAD Program Information](https://broadbandusa.ntia.doc.gov/funding-programs/broadband-equity-access-and-deployment-bead-program)
- [cori.data.fcc GitHub](https://github.com/ruralinnovation/cori.data.fcc)

---

**Last Updated:** December 15, 2025
**Data Coverage:** Census block level, all U.S. states
**Contact:** [data@ruralinnovation.us](mailto:data@ruralinnovation.us)
