---
title: "Census PEP Components of Change"
description: "Annual decomposition of population change into births, deaths, and migration"
author: "CORI Data Team"
date: "2026-02-02"
categories: ["Dataset", "Demographics"]
tags: ["Census", "Population", "Migration", "Vital Statistics", "County-Level"]
dataset:
  name: "Census Population Estimates Program - Components of Change"
  source: "U.S. Census Bureau"
  sourceUrl: "https://www.census.gov/data/tables/time-series/demo/popest/2020s-counties-detail.html"
  accessMethod: "Direct Download (FTP)"
  updateFrequency: "Annual"
  geographicLevel: "County, State, National"
  dataFormat: ["CSV"]
  featured: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Overview

The Census Population Estimates Program (PEP) Components of Change dataset decomposes annual population change into its underlying drivers:

- **Births** - Natural increase component
- **Deaths** - Natural decrease component
- **Migration** - Domestic and international migration

This decomposition reveals **why** counties are growing or declining, enabling targeted policy responses. For example:

- Is rural population decline due to out-migration or aging (more deaths than births)?
- Are amenity counties growing from migration or natural increase?
- How did pandemic mortality affect rural demographic trends?

::: callout-note
## Why Components Matter

- **Decomposed change** - Understand **why** population is changing
- **Administrative data** - Births and deaths from vital statistics (highly accurate)
- **Complete coverage** - All counties, all years
- **Faster than IRS migration** - Released ~12 months after reference year
:::

## Coverage

::: columns
::: {.column width="50%"}
**Geographic Coverage**

- County
- State
- National
:::

::: {.column width="50%"}
**Temporal Coverage**

- 2000-present (annual)
- CORI uses 2007-2023
- Released each March
- Universe: All U.S. counties
:::
:::

## Key Variables

### Natural Change

- **Births** - Total births in calendar year (from vital statistics)
- **Deaths** - Total deaths in calendar year (from vital statistics)
- **Natural Change** - Births minus deaths

### Migration

- **Domestic Migration** - Net migration from other U.S. counties
- **International Migration** - Net migration from other countries
- **Net Migration** - Domestic + international migration

### Residual

- **Residual** - Unexplained population change (census corrections, armed forces, group quarters)

### Total Population Change

- **Population Change** - Net change in population (natural change + net migration + residual)

## Access

::: panel-tabset
## Census FTP Downloads

Direct CSV downloads by vintage:

- [2020-2023 Components](https://www2.census.gov/programs-surveys/popest/tables/2020-2023/counties/totals/)
- [2010-2019 Components](https://www2.census.gov/programs-surveys/popest/tables/2010-2019/counties/totals/)
- [2000-2009 Components](https://www2.census.gov/programs-surveys/popest/tables/2000-2009/counties/totals/)

## CORI Projects

```{r}
#| eval: false

# From rural-economic-outlook project
# Load script: R/data_import/load_pep_components.R
# Combines multiple vintage files for 2007-2023 time series
```
:::

## Data Structure

**County Annual Format:**

- One row per county per year
- GEOID: 5-digit FIPS code
- Separate columns for each component

**Typical Variables:**

```
geoid, year, births, deaths, naturalchg,
domesticmig, internationalmig, netmig,
residual, population_change
```

## Methodology

### Component of Change Model

Population in year t is calculated as:

```
Pop(t) = Pop(t-1) + Births(t) - Deaths(t) + DomesticMig(t) + IntlMig(t) + Residual(t)
```

### Data Sources by Component

| Component | Primary Data Source |
|-----------|---------------------|
| **Births** | National Vital Statistics System (NVSS) - State birth certificates |
| **Deaths** | NVSS - State death certificates |
| **Domestic Migration** | **Residual estimation** - IRS tax records, Medicare enrollment, ACS |
| **International Migration** | DHS immigration data, ACS, emigration estimates |

::: callout-important
## Domestic Migration is Estimated

Domestic migration is **not directly measured**. It's estimated as a residual after accounting for births, deaths, and international migration.
:::

### Vintage Files

Census releases components in "vintage" batches:

- Each vintage may revise earlier years based on new census benchmarks
- CORI combines vintages to create consistent long time series
- Accepting most recent vintage for each year (no retroactive revisions)

## Use Cases

### In CORI Projects

| Project | Use Case |
|---------|----------|
| [Rural Economic Outlook](/projects/rural-economic-outlook/) | Natural increase vs. migration 2007-2023 |
| Demographic Narratives | Why are rural counties declining? |
| Pandemic Impact | Excess deaths 2020-2021 |
| Migration Analysis | Complement to [IRS migration data](/datasets/irs-migration/) |
| Aging Analysis | Natural decrease (deaths > births) in aging counties |

### Common Applications

- Understanding drivers of population change
- Aging and mortality analysis
- Migration pattern analysis
- Policy targeting (e.g., birth rates for maternal health, deaths for healthcare access)
- Scenario planning (what if birth rates increase?)

## Strengths & Limitations

::: columns
::: {.column width="50%"}
### Strengths

- Decomposed change (understand why)
- Administrative data (births/deaths accurate)
- Complete coverage
- Faster than IRS migration
- International migration included
:::

::: {.column width="50%"}
### Limitations

- Domestic migration is residual (estimated)
- No origin-destination flows
- No migration by income
- Group quarters challenge
- Vintage revisions
:::
:::

## Important Notes

### Natural Increase vs. Natural Decrease

::: panel-tabset
## Natural Increase

**Births > Deaths** (population growing from natural change)

## Natural Decrease

**Deaths > Births** (population declining from natural change)

As of 2023:

- **~50% of rural counties** have natural decrease
- Driven by aging population (fewer women of childbearing age, more elderly)
- Natural decrease is **accelerating** in rural America

Even if migration is positive, natural decrease can cause overall population decline.
:::

### Pandemic Mortality Impact (2020-2021)

Components data reveals COVID-19's demographic impact:

- **Excess deaths:** 2020-2021 death counts 15-20% above 2019 baseline
- Rural counties hit harder in 2021 (delta wave)
- Natural increase/decrease dramatically worsened

### Domestic Migration Estimation Challenges

Domestic migration is the **most uncertain** component:

- Estimated as residual after accounting for other components
- Subject to large revisions when census data released
- Can have negative migration even when county is attractive (if births + international migration undercounted)

**Best practice:** Use [IRS migration data](/datasets/irs-migration/) for detailed migration analysis (county-to-county flows, income).

### Vintage Reconciliation Strategy

::: callout-tip
## CORI's Approach

CORI's `load_pep_components.R` script combines 3 vintage files:

1. 2000-2009 vintage (accepted as-is)
2. 2010-2019 vintage (accepted as-is)
3. 2020-2023 vintage (most recent)

**No retroactive revisions** - Accepting each vintage's estimates for its period prevents time series discontinuities.
:::

### Additive Components

All components are **additive**:

- Births, deaths, domestic migration, international migration, residual **sum** to total population change
- Enables accounting for 100% of population change
- Rates can be calculated using same denominator (population)

## Related Datasets

- [Census Population Estimates](/datasets/census-population-estimates/) - Total population (numerator for rates)
- [IRS County-to-County Migration](/datasets/irs-migration/) - Detailed migration flows and income (complementary)
- [National Vital Statistics System (NVSS)](https://www.cdc.gov/nchs/nvss/) - Direct access to birth/death microdata

## R Packages

### Recommended Packages

```{r}
#| eval: false

# Install packages
install.packages("tidycensus")   # Can access components via get_estimates()
```

### Usage Example

```{r}
#| eval: false

library(tidycensus)

# Get components of change
components <- get_estimates(
  geography = "county",
  product = "components",
  year = 2023,
  vintage = 2023
)
```

## Additional Resources

- [Census PEP Methodology](https://www.census.gov/programs-surveys/popest/technical-documentation/methodology.html)
- [Rural Economic Outlook Project](https://github.com/ruralinnovation/data-rural-economic-outlook)

---

**Last Updated:** February 2, 2026
**Data Coverage:** 2000-2023 (CORI uses 2007-2023)
**Contact:** [data@ruralinnovation.us](mailto:data@ruralinnovation.us)
