---
title: "ruraldefinitions"
description: "R package for accessing federal rural definitions at multiple geographic levels"
author: "Camden Blatchly"
date: "2026-02-02"
categories: ["R Package", "Rural Classification"]
tags: ["USDA", "Census", "RUCA", "Rural Definitions", "Geography"]
package:
  name: "ruraldefinitions"
  githubUrl: "https://github.com/ruralinnovation/ruraldefinitions"
  installCommand: 'devtools::install_github("ruralinnovation/ruraldefinitions")'
  status: "stable"
  version: "0.1.0"
  maintainer: "Camden Blatchly"
  featured: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Overview

The `ruraldefinitions` package provides easy access to federal rural definitions from multiple U.S. agencies. It aggregates seven major rural classification systems, making them readily accessible to R programmers for comparative and historical analysis.

This package is essential for researchers and analysts who need consistent, well-documented rural definitions across different federal agencies' methodologies and time periods.

::: callout-note
## Why Use This Package

- **One-stop access** - All major federal rural definitions in one place
- **Standardized format** - Consistent column names and structure across definitions
- **Historical depth** - 50+ years of rural classifications (1974-2024)
- **Multiple geographies** - County, tract, and block-level classifications
- **Expert guidance** - Built-in recommendations for definition selection
:::

## Installation

::: panel-tabset
## GitHub

```{r}
#| eval: false

# Install devtools if you haven't already
install.packages("devtools")

# Install ruraldefinitions
devtools::install_github("ruralinnovation/ruraldefinitions")
```

## AWS S3 Access

For non-R users, CSV files are available directly:

- [Download from S3](https://ruraldefinitions.s3.us-east-1.amazonaws.com)
:::

## Quick Start

```{r}
#| eval: false

library(ruraldefinitions)

# Access pre-loaded county-level definitions
head(cbsa_2020)  # Core-Based Statistical Areas
head(rucc_2023)  # Rural-Urban Continuum Codes

# Download tract-level definitions from S3
cori_tracts <- get_definition("cori", 2020)
ruca_tracts <- get_definition("ruca", 2010)

# All definitions have standardized columns:
# - geoid: Geographic identifier
# - name: Definition name
# - year: Year released
# - rural_def: Detailed classification
# - is_rural: "Rural" or "Nonrural"
```

## Available Definitions

The package provides access to **7 major federal rural definitions**:

| Definition | Type | Years Available | Geographic Level |
|-----------|------|-----------------|-----------------|
| **CORI** | Custom hybrid definition | 2020 | Tract |
| **RUCA** | Rural-Urban Commuting Area | 1990, 2000, 2010 | Tract |
| **CBSA** | Core-Based Statistical Areas | 2020, 2023 | County |
| **NCHS** | National Center for Health Statistics | 1990, 2006, 2013, 2023 | County |
| **RUCC** | Rural-Urban Continuum Codes | 1974, 1983, 1993, 2003, 2013, 2023 | County |
| **UIC** | Urban Influence Codes | 1993, 2003, 2013, 2024 | County |
| **Census** | Census Bureau | 2010, 2020 | Block |

## Key Functions

### Data Access

```{r}
#| eval: false

# Download tract-level and block-level definitions from AWS S3
get_definition(
  name = "cori",     # Definition type
  year = 2020        # Year of classification
)

# Pre-loaded county-level definitions (no download needed)
data(cbsa_2020)
data(nchs_2023)
data(rucc_2023)
data(uic_2024)
```

## Expert Recommendations

::: callout-tip
## Choosing the Right Definition

The package includes guidance on definition selection based on research by rural innovation experts:

**For county-level analysis:**
- Use **CBSA** or "Nonmetro" definitions
- Best for state/regional comparisons
- Aligns with OMB standards

**For tract-level analysis:**
- Use **CORI** definition (hybrid of Nonmetro + RUCA)
- Captures rural tracts within metropolitan counties
- Better for granular analysis

These recommendations align with the Urban Institute's "Do No Harm Guide" for equitable data narratives.
:::

## Common Use Cases

### Basic Rural Classification

```{r}
#| eval: false

library(dplyr)

# Get CORI rural definition
rural_tracts <- get_definition("cori", 2020)

# Filter to rural areas only
rural_only <- rural_tracts |>
  filter(is_rural == "Rural")

# Count by classification
rural_tracts |>
  count(rural_def, is_rural)
```

### Comparing Definitions

```{r}
#| eval: false

# Compare RUCA and CORI definitions
ruca_2010 <- get_definition("ruca", 2010)
cori_2020 <- get_definition("cori", 2020)

# Join and compare
comparison <- ruca_2010 |>
  left_join(cori_2020, by = "geoid", suffix = c("_ruca", "_cori")) |>
  count(is_rural_ruca, is_rural_cori)
```

### Historical Analysis

```{r}
#| eval: false

# Track rural classification changes over time
rucc_1993 <- rucc_1993  # Pre-loaded data
rucc_2003 <- rucc_2003
rucc_2013 <- rucc_2013
rucc_2023 <- rucc_2023

# Join across years
historical <- rucc_2023 |>
  select(geoid, rural_def_2023 = rural_def) |>
  left_join(
    rucc_2013 |> select(geoid, rural_def_2013 = rural_def),
    by = "geoid"
  ) |>
  left_join(
    rucc_2003 |> select(geoid, rural_def_2003 = rural_def),
    by = "geoid"
  )

# Identify counties that changed classification
changed <- historical |>
  filter(rural_def_2023 != rural_def_2003)
```

### Integration with Other Data

```{r}
#| eval: false

library(tidycensus)

# Get Census population data
pop_data <- get_acs(
  geography = "tract",
  variables = "B01001_001",
  state = "VT",
  year = 2020
)

# Add rural classification
pop_rural <- pop_data |>
  left_join(
    get_definition("cori", 2020),
    by = c("GEOID" = "geoid")
  )

# Calculate rural vs. urban population
pop_rural |>
  group_by(is_rural) |>
  summarize(
    total_population = sum(estimate, na.rm = TRUE)
  )
```

## Data Structure

**Standardized Format:**

All definitions share a common structure:

```
geoid       - Geographic identifier (FIPS code)
name        - Definition name (e.g., "CORI", "RUCA")
year        - Year of classification
rural_def   - Detailed classification (e.g., "Metropolitan", "Micropolitan")
is_rural    - Binary: "Rural" or "Nonrural"
```

**Example:**

| geoid | name | year | rural_def | is_rural |
|-------|------|------|-----------|----------|
| 50001020100 | CORI | 2020 | Rural | Rural |
| 50001020200 | CORI | 2020 | Metropolitan | Nonrural |

## Definition Details

### CORI Definition

Custom hybrid combining:
- **Nonmetro counties** (CBSA classification)
- **RUCA codes 4-10** (rural and small town tracts)

Captures rural areas more accurately than county-level definitions alone.

### RUCA (Rural-Urban Commuting Area)

Census tract-level classification based on:
- Population density
- Urbanization
- Daily commuting patterns

10 primary codes, with secondary codes for finer distinctions.

### CBSA (Core-Based Statistical Areas)

OMB delineations of metro and micro areas:
- **Metropolitan**: 50,000+ population urban core
- **Micropolitan**: 10,000-49,999 population core
- **Rural/Noncore**: Neither metro nor micro

### RUCC (Rural-Urban Continuum Codes)

USDA 9-category classification:
- Codes 1-3: Metropolitan
- Codes 4-9: Nonmetropolitan (varying levels of urbanization and adjacency)

### NCHS (National Center for Health Statistics)

6-category urban-rural classification:
- Large central metro
- Large fringe metro
- Medium metro
- Small metro
- Micropolitan
- Noncore

### UIC (Urban Influence Codes)

USDA 12-category classification based on:
- Metro/nonmetro status
- Population size
- Adjacency to metro areas

### Census Urban/Rural

Census Bureau's official urban-rural classification:
- Based on population density and land use
- Block-level precision

## Performance Considerations

::: callout-tip
## Loading Strategy

- **County data** - Pre-loaded, instant access
- **Tract data** - Downloads from S3 (~1-2 MB)
- **Block data** - Large files, use selectively

Cache downloaded data locally to avoid repeated S3 calls.
:::

## Data Hosting

- **County-level data:** Included in the package for immediate access
- **Tract and block-level data:** Hosted on AWS S3 and downloaded on-demand
- **CSV Downloads:** Available directly from [AWS S3](https://ruraldefinitions.s3.us-east-1.amazonaws.com) for non-R users

## Use Cases at CORI

This package is used across CORI projects for:

- Consistent geographic classifications
- Rural vs. urban comparisons
- Equity analysis
- Historical trend analysis
- Spatial data enrichment

## Related Resources

### Internal Resources

- CORI research on defining rural America
- Geographic analysis projects
- Equity and access studies

### External Links

- [USDA ERS Rural Definitions](https://www.ers.usda.gov/topics/rural-economy-population/rural-classifications/)
- [Census Urban/Rural Classification](https://www.census.gov/programs-surveys/geography/guidance/geo-areas/urban-rural.html)
- [RUCA Data](https://www.ers.usda.gov/data-products/rural-urban-commuting-area-codes/)
- [NCHS Urban-Rural Classification](https://www.cdc.gov/nchs/data_access/urban_rural.htm)

## Key Features

1. **Comprehensive Coverage** - Aggregates 7 different federal rural definitions spanning 50+ years (1974-2024)
2. **Flexible Geographic Levels** - Supports county, tract, and block-level analysis
3. **Standardized Format** - All definitions use consistent column names and structure
4. **Easy Access** - Pre-loaded county data; on-demand loading for detailed geographies
5. **Historical Analysis** - Multiple years available for temporal comparisons
6. **Open License** - MIT licensed for broad reuse

## Documentation

- [GitHub Repository](https://github.com/ruralinnovation/ruraldefinitions)
- [README and User Guide](https://github.com/ruralinnovation/ruraldefinitions#readme)
- [Data Dictionary](https://github.com/ruralinnovation/ruraldefinitions/wiki/Data-Dictionary)

---

**Contact:** Camden Blatchly

For bug reports or feature requests, please open an issue on [GitHub](https://github.com/ruralinnovation/ruraldefinitions/issues).
