---
title: "BLS Quarterly Census of Employment and Wages (QCEW)"
description: "Comprehensive county-level employment and wage data by industry from the Bureau of Labor Statistics"
author: "CORI Data Team"
date: "2026-02-02"
categories: ["Dataset", "Labor Market"]
tags: ["BLS", "Employment", "Wages", "Industry", "County-Level"]
dataset:
  name: "Quarterly Census of Employment and Wages (QCEW)"
  source: "Bureau of Labor Statistics, U.S. Department of Labor"
  sourceUrl: "https://www.bls.gov/cew/"
  accessMethod: "BLS API and Bulk Download"
  updateFrequency: "Quarterly (with 6-month lag)"
  geographicLevel: "County, MSA, State, National"
  dataFormat: ["CSV", "API", "Flat Files"]
  featured: true
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Overview

The Quarterly Census of Employment and Wages (QCEW) is a comprehensive
dataset covering 95% of U.S. jobs. Unlike survey-based employment data,
QCEW is a **census** based on mandatory quarterly unemployment insurance
(UI) tax filings from employers.

QCEW provides employment counts and wage data at detailed industry
levels (6-digit NAICS codes) for every county in the U.S., making it the
gold standard for understanding local labor markets and economic
structure.

::: callout-note
## Why QCEW is Essential

-   **Census data** - Not a survey, covers 95% of U.S. jobs
-   **Industry detail** - 6-digit NAICS codes (1,000+ industries)
-   **Historical depth** - Consistent data back to 1990
:::

## Coverage

::: columns
::: {.column width="50%"}
**Geographic Coverage** - County level - Metropolitan/Micropolitan
Areas - State level - National totals
:::

::: {.column width="50%"}
**Temporal Coverage** - 1990-present (quarterly and annual) - Quarterly
updates (\~6-month lag) - Annual averages (recommended for trends)
:::
:::

**Universe:** 95% of U.S. jobs (excludes self-employed, farm workers,
household workers)

## Key Variables

### Employment Metrics

```{r}
#| echo: false
#| label: tbl-employment-vars
#| tbl-cap: "Key Employment Variables in QCEW"

library(tibble)
library(knitr)

tribble(
  ~Variable, ~Description, ~Use_Case,
  "Annual Average Employment", "Average monthly employment across 4 quarters", "Trend analysis",
  "Monthly Employment Levels", "Employment for each of 3 months per quarter", "Seasonal patterns",
  "Establishment Count", "Number of business establishments", "Business density",
  "Employment by Industry", "Detailed NAICS breakdowns", "Sectoral analysis"
) |> kable()
```

### Wage Metrics

-   **Total Wages** - Sum of quarterly wages paid
-   **Average Annual Pay** - Total wages / annual average employment
-   **Average Weekly Wage** - Quarterly wages / (13 weeks × employment)

### Derived Metrics (CORI Projects)

::: panel-tabset
## Real Wages

Nominal wages deflated to **2023 dollars** using BLS CPI-U-RS:

```{r}
#| eval: false

# Example deflation calculation
real_wage_2023 = nominal_wage * (cpi_2023 / cpi_year)
```

## Industry Shares

Employment shares by sector (% of total employment)

## HHI Index

Herfindahl-Hirschman Index measuring industry concentration (0-10,000
scale):

```{r}
#| eval: false

# HHI calculation
HHI = sum((employment_share_i)^2) * 10000
```

-   HHI \< 1,500: Unconcentrated economy
-   HHI 1,500-2,500: Moderate concentration
-   HHI \> 2,500: High concentration

## Economic Base

-   **Tradable Goods** - Manufacturing, mining, agriculture
-   **Tradable Services** - Information, finance, professional services
-   **Local Services** - Retail, healthcare, hospitality
:::

## Access

### Using R

::: panel-tabset
## blsAPI

```{r}
#| eval: false

library(blsAPI)

# Get employment for a specific county
series_id <- "ENU0601510510"  # Example series ID
qcew_data <- blsAPI(series_id, startyear = 2010, endyear = 2024)
```

## Bulk Download

```{r}
#| eval: false

# Download annual averages CSV
url <- "https://www.bls.gov/cew/downloadable-data-files.htm"
download.file(url, "qcew_annual.csv")
```

## CORI Method

```{r}
#| eval: false

# From rural-economic-outlook project
# Load script: R/data_import/load_qcew.R
source("R/data_import/load_qcew.R")

# Includes wage deflation and HHI calculation
qcew_data <- load_qcew(years = 2007:2024)
```
:::

### Interactive Data Viewer

Visit the [QCEW Data
Viewer](https://data.bls.gov/cew/apps/data_views/data_views.htm) for
interactive queries.

## Methodology

### Data Collection

QCEW is built from three sources:

```{mermaid}
%%| fig-width: 6
%%| fig-height: 3

graph LR
    A[UI-Covered Employers<br/>Quarterly UI-3 Forms] --> D[QCEW Database]
    B[Federal Agencies<br/>USPS, Military] --> D
    C[Railroad Retirement Board<br/>Railroad Workers] --> D
    D --> E[95% of Civilian<br/>Employment]
```

::: callout-warning
## Excluded from QCEW

-   Self-employed individuals (no employees)
-   Farm workers (some states)
-   Household workers (domestic employees)
-   Unincorporated businesses
:::

### Wage Deflation (CORI Method)

Nominal wages converted to real 2023 dollars using **BLS CPI-U-RS**:

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show deflation function"

deflate_wages <- function(nominal_wage, year, base_year = 2023) {
  # CPI-U-RS values (example)
  cpi_values <- c(
    "2020" = 306.8,
    "2021" = 319.4,
    "2022" = 348.0,
    "2023" = 360.0
  )

  real_wage <- nominal_wage * (cpi_values[as.character(base_year)] /
                                cpi_values[as.character(year)])
  return(real_wage)
}
```

**Why CPI-U-RS?** More consistent for historical comparisons than
standard CPI-U.

## Use Cases

### In CORI Projects

| Project                                                     | Use Case                                               |
|----------------------------------|--------------------------------------|
| [Rural Economic Outlook](/projects/rural-economic-outlook/) | Employment growth, wage trends 2007-2024               |
| Broadband Equity                                            | Tech sector employment by county                       |
| Industry Analysis                                           | Sectoral employment shares and concentration           |
| Wage Gap Analysis                                           | Rural vs. metro pay differentials (inflation-adjusted) |

: {tbl-colwidths="\[30,70\]"}

### Common Applications

-   Labor market analysis
-   Economic development targeting
-   Industry cluster identification
-   Wage competitiveness studies
-   Commuting pattern analysis (with LODES)

## Strengths & Limitations

::: columns
::: {.column width="50%"}
### ✅ Strengths

-   Census, not survey (no sampling error)
-   95% job coverage
-   6-digit NAICS industry detail
-   Historical depth (1990+)
-   Quarterly frequency
-   Official data source
:::

::: {.column width="50%"}
### ⚠️ Limitations

-   6-month lag in release
-   Excludes self-employment
-   Disclosure suppression (\<3 establishments)
-   Location of work, not residence
-   Nominal wages (need deflation)
-   Quarterly volatility
:::
:::

## Important Notes

### Disclosure Rules

::: callout-important
## Privacy Protection

To protect business confidentiality: - Industries with **fewer than 3
establishments suppressed** - Rural counties often have more suppression
than urban - **Solution:** Use 2-digit NAICS for consistent county
coverage
:::

### BEA Table Replacement

As of 2023, QCEW is the **recommended source** for county employment
after 2022:

| BEA Table                          | Status                     | Replacement |
|-------------------------|-------------------|----------------------------|
| CAEMP25N (Employment by industry)  | ❌ Discontinued after 2022 | ✅ QCEW     |
| CAINC6N (Compensation by industry) | ❌ Discontinued after 2022 | ✅ QCEW     |

QCEW provides comparable industry breakdowns with more frequent updates.

### Annual vs. Quarterly Data

```{r}
#| echo: false
#| label: fig-annual-vs-quarterly
#| fig-cap: "When to Use Annual vs. Quarterly Data"
#| fig-width: 8
#| fig-height: 3

library(ggplot2)
library(dplyr)

tibble(
  x = c(1, 1, 2, 2),
  y = c(1, 2, 1, 2),
  label = c("Annual Averages", "Quarterly Data",
            "✓ Trend analysis\n✓ Smooths seasonality\n✓ Year-over-year comparisons",
            "✓ Recession timing\n✓ Seasonal industries\n✓ Real-time monitoring")
) |>
  ggplot(aes(x, y, label = label)) +
  geom_text(size = 4, hjust = 0.5, vjust = 0.5) +
  theme_void() +
  coord_cartesian(xlim = c(0.5, 2.5), ylim = c(0.5, 2.5))
```

## Related Datasets

-   [BEA Real GDP](/datasets/bea-real-gdp/) - Economic output (GDP) to
    complement employment
-   [Census Business Dynamics Statistics](/datasets/census-bds/) -
    Business entry/exit rates
-   [Census Business Formation Statistics](/datasets/census-bfs/) -
    Leading indicator of employment
-   [American Community Survey](/datasets/american-community-survey/) -
    Occupation detail

## R Packages

### Recommended Packages

```{r}
#| eval: false

# Install packages
install.packages("blsAPI")        # BLS API wrapper
install.packages("blscrapeR")     # Bulk download tools
```

### CORI Packages

For CORI internal use:

```{r}
#| eval: false

# CORI data infrastructure
library(coriverse)
qcew_data <- get_qcew(geography = "county", years = 2007:2024)
```

## Additional Resources

-   [QCEW Overview &
    Handbook](https://www.bls.gov/opub/hom/cew/home.htm)
-   [BLS Data
    Tools](https://data.bls.gov/cew/apps/data_views/data_views.htm)
-   [Rural Economic Outlook
    Project](https://github.com/ruralinnovation/data-rural-economic-outlook)

------------------------------------------------------------------------

**Last Updated:** February 2, 2026 **Data Coverage:** 1990-2024
(quarterly), 2007-2024 (CORI focus) **Contact:**
[data\@ruralinnovation.us](mailto:data@ruralinnovation.us)
