---
title: "Census Business Formation Statistics (BFS)"
description: "Leading indicator of new business activity based on IRS business applications"
author: "CORI Data Team"
date: "2026-02-02"
categories: ["Dataset", "Entrepreneurship"]
tags: ["Census", "Business Formation", "Startups", "Leading Indicator"]
dataset:
  name: "Business Formation Statistics (BFS)"
  source: "U.S. Census Bureau"
  sourceUrl: "https://www.census.gov/econ/bfs/"
  accessMethod: "Census API and Bulk Download"
  updateFrequency: "Monthly (state/national), Annual (county)"
  geographicLevel: "County, State, National"
  dataFormat: ["CSV", "API", "Excel"]
  featured: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Overview

The Business Formation Statistics (BFS) provides high-frequency data on new business applications based on IRS Employer Identification Number (EIN) applications (Form SS-4). BFS serves as a **leading indicator** of entrepreneurial activity and future business formation.

BFS includes two key series:

- **Business Applications (BA)** - All applications likely to be a new business
- **High-Propensity Business Applications (HBA)** - Applications with high likelihood of becoming employer businesses

::: callout-note
## Why BFS is Essential

- **High frequency** - Monthly updates (state/national) for real-time tracking
- **Leading indicator** - Predicts future business formation
- **Universal coverage** - Administrative data, not survey
- **Long time series** - Data back to 2004
:::

## Coverage

::: columns
::: {.column width="50%"}
**Geographic Coverage**

- County (annual, BA only)
- State (monthly and annual, BA and HBA)
- National (monthly and annual, BA and HBA)
:::

::: {.column width="50%"}
**Temporal Coverage**

- 2004-present
- Monthly (state/national, ~2-week lag)
- Annual (county, ~4-month lag)
- Universe: IRS Form SS-4 applications
:::
:::

## Key Variables

### Business Applications (BA)

- **Total Applications** - All filtered business applications
- **Applications by Sector** - NAICS 2-digit industry classification
- **Applications by Organizational Type** - Corporation, partnership, LLC, etc.

### High-Propensity Business Applications (HBA)

- **High-Propensity Applications** - Subset of BA with high likelihood of hiring employees
- **HBA by Sector** - Industry classification
- **Not available at county level** - Only state/national

### Derived Metrics

- **Applications per 1,000 population** - Normalized startup rate
- **Year-over-year growth rate** - Business formation momentum

## Access

::: panel-tabset
## Census BFS Data Viewer

Interactive tool:

- [BFS Data Viewer](https://www.census.gov/econ/bfs/data.html)

## Bulk Download

Annual county files:

- [BFS Datasets](https://www.census.gov/econ/bfs/data.html)

## CORI Projects

```{r}
#| eval: false

# From rural-economic-outlook project
# Load script: R/data_import/load_bfs.R
# Currently pulls BA series (HBA not available at county level)
```
:::

## Data Structure

**County Annual Format:**

- One row per county per year
- GEOID: 5-digit FIPS code
- BA series only (no HBA at county level)

**State Monthly Format:**

- One row per state per month
- Includes both BA and HBA series

**Typical Variables:**

```
geoid, year, business_applications,
applications_per_1k_population
```

## Methodology

### Data Source: IRS Form SS-4

BFS is derived from IRS applications for Employer Identification Numbers (EINs):

- Required for hiring employees, opening bank accounts, filing taxes
- Filed before business operations begin
- Indicator of intent to start business

### Filtering Algorithm

Census filters out non-business applications:

- Personal use EINs
- Trusts and estates
- Government entities
- Non-operating entities

Result: **Business Applications (BA)** series

### High-Propensity Business Applications (HBA)

HBA subset includes applications with **high likelihood of becoming employer firms**:

- Specific organizational structures (corporations, partnerships, LLCs)
- Wage payment intentions
- Certain industry classifications

::: callout-important
## Conversion Rates

**From NBER research (2024):**

- ~33% of BA are classified as HBA
- ~34.5% of HBA become employer firms within 18 months
- **Net result:** ~11-12% of all BA become employer businesses
:::

### Leading Indicator Properties

BFS is a **leading indicator** because:

- Applications filed **before** business operations begin
- Average lag: **18 months** from application to business operation
- Correlation with future BDS entry rates: 0.7-0.8

Research validation:

- Bayard et al. (2024, NBER) confirms HBA predicts employer firm formation
- HBA strongly correlated with subsequent job creation

## Use Cases

### In CORI Projects

| Project | Use Case |
|---------|----------|
| [Rural Economic Outlook](/projects/rural-economic-outlook/) | Business formation trends 2007-2024 |
| Leading Indicator Analysis | Early signal of entrepreneurship momentum |
| Pandemic Response | Tracked surge in applications during 2020-2021 |
| Sectoral Analysis | Which industries seeing most new applications |

### Common Applications

- Nowcasting economic recovery
- Entrepreneurship ecosystem assessment
- Policy evaluation (e.g., pandemic stimulus impact)
- Regional competitiveness tracking

## Strengths & Limitations

::: columns
::: {.column width="50%"}
### Strengths

- High frequency (monthly)
- Leading indicator
- Universal coverage
- Long time series (2004+)
- Industry detail
:::

::: {.column width="50%"}
### Limitations

- County: BA only (no HBA)
- County: Annual only (no monthly)
- Applications ≠ businesses (~11-12% become employers)
- Organizational structure changes
- No employment data
- Disclosure suppression
:::
:::

## Important Notes

### County-Level Limitations

::: callout-warning
## Critical County Data Limitations

**County BFS data has significant limitations:**

- **Only Business Applications (BA)** available, not HBA
- Only ~33% of BA are high-propensity (HBA)
- Only ~11-12% of BA actually become employer firms
- **18-month lag** on average from application to operation

**Interpretation:** County BA counts should be interpreted as **intent to start business**, not actual business formation. For actual employer businesses, use [Census BDS](/datasets/census-bds/).
:::

### Recent Methodology Changes (January 2026)

Census revised HBA methodology:

- **Internet sales now excluded** from HBA/CBA series
- Reduces HBA/CBA counts going forward
- Creates time series break at January 2026
- Documentation: [BFS Methodology Update](https://www.census.gov/econ/bfs/methodology.html)

### HBA vs. BA

| Metric | BA | HBA |
|--------|----|----|
| **Definition** | All filtered business applications | Subset with high likelihood of hiring |
| **County data** | ✅ Yes (annual) | ❌ No |
| **Monthly data** | ✅ Yes (state/national) | ✅ Yes (state/national) |
| **Conversion to employers** | ~11-12% | ~34.5% |
| **Use case** | Broad entrepreneurial intent | Serious business formation |

### Link to BDS

BFS is a **leading indicator** of [Business Dynamics Statistics (BDS)](/datasets/census-bds/):

```
BFS (application) → 18 months → BDS (employer firm)
```

For **historical analysis**, use BDS (actual firms). For **nowcasting**, use BFS (applications).

## Detailed Documentation

For comprehensive technical details, see:

- [BFS Data Source Profile](/R/data_import/BFS_DATA_SOURCE_PROFILE.md) in rural-economic-outlook project
- Conversion funnel documentation
- Leading indicator validation research (Bayard et al., 2024)

## Related Datasets

- [Census Business Dynamics Statistics](/datasets/census-bds/) - Actual employer firm formation (follows BFS by 18 months)
- [Census Nonemployer Statistics](/datasets/census-nonemployer/) - Self-employed businesses
- [QCEW Employment](/datasets/qcew-employment-wages/) - Employment at established firms

## R Packages

### Recommended Packages

```{r}
#| eval: false

# Install packages
install.packages("censusapi")    # General Census API wrapper
```

### Usage Example

```{r}
#| eval: false

library(censusapi)

# Note: BFS API access varies by series
# Check Census documentation for current endpoints
```

## Additional Resources

- [BFS Methodology](https://www.census.gov/econ/bfs/methodology.html)
- [Bayard et al. (2024) NBER Paper](https://www.nber.org/papers/w32000) - HBA as leading indicator
- [Rural Economic Outlook Project](https://github.com/ruralinnovation/data-rural-economic-outlook)

---

**Last Updated:** February 2, 2026
**Data Coverage:** 2004-2024 (county annual from 2007)
**Contact:** [data@ruralinnovation.us](mailto:data@ruralinnovation.us)
