---
title: "IRS County-to-County Migration Data"
description: "Annual migration flows and income data based on tax return addresses"
author: "CORI Data Team"
date: "2026-02-02"
categories: ["Dataset", "Migration"]
tags: ["IRS", "Migration", "Population", "County-Level"]
dataset:
  name: "IRS Statistics of Income (SOI) County-to-County Migration"
  source: "Internal Revenue Service"
  sourceUrl: "https://www.irs.gov/statistics/soi-tax-stats-migration-data"
  accessMethod: "Manual Download"
  updateFrequency: "Annual"
  geographicLevel: "County-to-County Flows"
  dataFormat: ["CSV", "Excel"]
  featured: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Overview

The IRS Statistics of Income (SOI) County-to-County Migration Data tracks the movement of people and income across U.S. counties based on year-to-year changes in tax return filing addresses. This dataset provides the most comprehensive view of **where people are moving** and **how much income they bring with them**.

Unlike Census migration data (which is survey-based), IRS migration data is based on **administrative tax records**, covering the vast majority of the U.S. population.

::: callout-note
## Why IRS Migration Data is Essential

- **Administrative data** - Based on tax records, not surveys (high accuracy)
- **Income flows** - Unique ability to track wealth movement with people
- **Origin-destination detail** - Know exactly where people are moving from/to
- **High coverage** - ~95% of population (tax filers + dependents)
- **Long time series** - Data back to 1991
:::

## Coverage

::: columns
::: {.column width="50%"}
**Geographic Coverage**

- County-to-county migration flows
- Origin-destination pairs
:::

::: {.column width="50%"}
**Temporal Coverage**

- 1991-present (annual)
- CORI uses 2007-2022
- Released ~2 years after tax year
- Universe: All tax filers (~95% of U.S.)
:::
:::

## Key Variables

### Migration Flows

- **Returns (In-flow)** - Number of tax returns that moved into the county
- **Returns (Out-flow)** - Number of tax returns that moved out of the county
- **Exemptions (In-flow)** - Number of people who moved in (returns + dependents)
- **Exemptions (Out-flow)** - Number of people who moved out
- **Net Migration** - In-flow minus out-flow

### Income Flows

- **Adjusted Gross Income (AGI) In-flow** - Total AGI of in-migrants (thousands of dollars)
- **AGI Out-flow** - Total AGI of out-migrants
- **Net AGI Flow** - In-flow minus out-flow

### Derived Metrics (CORI Projects)

- **Net Migration Rate** - Net migration as % of county population
- **Net AGI Migration Rate** - Net AGI flow as % of county total AGI
- **Net Migration by Income Bracket** - Who is moving? (high/middle/low income)

## Access

::: panel-tabset
## IRS Website

Manual download of annual CSV files:

- [IRS SOI Migration Data](https://www.irs.gov/statistics/soi-tax-stats-migration-data)

Files organized by tax year (e.g., "1920.csv" for 2019-2020 migration).

## CORI Projects

```{r}
#| eval: false

# From rural-economic-outlook project
# Load script: R/data_import/load_irs_soi.R
# Aggregates county-pair flows to net county totals
```
:::

## Data Structure

### Raw Format (County-to-County Pairs)

- One row per origin-destination county pair
- Sparse matrix (only pairs with ≥10 returns reported)
- FIPS codes for origin (y1) and destination (y2)

**Example:**

```
origin_fips, dest_fips, returns, exemptions, agi_thousands
06001, 06013, 245, 512, 18750
06001, 36061, 89, 156, 9820
```

### Aggregated Format (Net Flows by County)

CORI aggregates to net flows:

```
geoid, year, net_migration_count, net_agi_flow,
share_of_net_migration, share_of_net_agi_migration
```

## Methodology

### Data Source: Tax Returns

IRS tracks address changes on Form 1040:

- Year 1 (y1) address vs. Year 2 (y2) address
- If addresses differ and in different counties → migration
- Returns matched using Social Security Numbers

**Coverage:**

- **Tax filers** - Required to file if income above threshold
- **Dependents** - Counted as "exemptions"
- **Non-filers excluded** - Very low-income, undocumented immigrants

### Privacy Protection

To protect taxpayer privacy:

- **Flows < 10 returns suppressed** - Small flows not reported
- **Aggregated only** - No individual-level data
- **Same-county moves excluded** - Only inter-county migration

### AGI as Income Proxy

Adjusted Gross Income (AGI) measures:

::: columns
::: {.column width="50%"}
**Included:**

- Wages and salaries
- Business income
- Investment income
- Retirement income
:::

::: {.column width="50%"}
**Excluded:**

- Social Security (reported separately)
- Tax-exempt interest
- Unrealized capital gains
:::
:::

**AGI reflects "movable" income** but has limitations.

## Use Cases

### In CORI Projects

| Project | Use Case |
|---------|----------|
| [Rural Economic Outlook](/projects/rural-economic-outlook/) | Migration trends 2007-2022 |
| Brain Drain Analysis | Are high-income households leaving rural areas? |
| Economic Impact | Income flows associated with migration |
| Regional Competitiveness | Which counties are gaining/losing residents and wealth? |

### Common Applications

- Regional economic development strategy
- Understanding pandemic migration (2020-2021)
- Housing demand forecasting
- Tax revenue projections
- Workforce planning

## Strengths & Limitations

::: columns
::: {.column width="50%"}
### Strengths

- Administrative data (high accuracy)
- Income flows (unique capability)
- Origin-destination detail
- High coverage (~95%)
- Long time series (1991+)
:::

::: {.column width="50%"}
### Limitations

- Lagged release (18-24 months)
- Excludes non-filers
- Tax year vs. calendar year
- Suppression (<10 returns)
- Address changes only
- AGI not deflated (nominal dollars)
:::
:::

## Important Notes

### Net Migration vs. Gross Flows

IRS data includes both:

- **Gross in-flow** - Total people moving in
- **Gross out-flow** - Total people moving out
- **Net migration** - In-flow minus out-flow

::: callout-tip
## Example

County gains 500 in-migrants, loses 400 out-migrants

- **Gross in-flow:** 500
- **Net migration:** +100

Many rural counties have **high gross flows** (people moving in and out) but **small net migration** (in and out roughly balance).
:::

### AGI Deflation (Important!)

IRS reports AGI in **nominal dollars** (not adjusted for inflation). For time series analysis:

```{r}
#| eval: false

# Deflate to real 2023 dollars using CPI-U-RS
real_agi_2023 = nominal_agi * (cpi_2023 / cpi_year)
```

CORI's `load_irs_soi.R` script should include deflation (currently a TODO item).

### Pandemic Migration Surge (2020-2021)

IRS data shows unprecedented migration during COVID-19:

- 2020-2021 migration rates 30-50% higher than 2019
- Large cities lost population to suburbs and smaller metros
- Rural recreation counties (mountains, coasts) gained
- Remote work enabled moves

Rural net migration turned **positive** for first time in decades in many counties.

### Relationship to Census Data

IRS migration complements [Census PEP Components of Change](/datasets/census-pep-components/):

| Dataset | Strengths | Limitations |
|---------|-----------|-------------|
| **IRS** | Origin-destination flows, income data, administrative | 18-month lag, excludes non-filers |
| **Census Components** | Faster release, all population, separate domestic/international | No origin-destination, no income |

**Best practice:** Use IRS for detailed flow analysis; use Census Components for timely total migration.

## Related Datasets

- [Census PEP Components of Change](/datasets/census-pep-components/) - Faster migration counts (no income or flows)
- [Census Population Estimates](/datasets/census-population-estimates/) - Population for per-capita rates
- [American Community Survey](/datasets/american-community-survey/) - Migration within past year (survey)

## R Packages

### Custom Scripts

```{r}
#| eval: false

# No dedicated R package available
# CORI uses custom scripts for processing

# From rural-economic-outlook project
source("R/data_import/load_irs_soi.R")
```

## Additional Resources

- [IRS SOI Documentation](https://www.irs.gov/statistics/soi-tax-stats-migration-data-methodology)
- [Rural Economic Outlook Project](https://github.com/ruralinnovation/data-rural-economic-outlook)

---

**Last Updated:** February 2, 2026
**Data Coverage:** 1991-2022 (CORI uses 2007-2022)
**Contact:** [data@ruralinnovation.us](mailto:data@ruralinnovation.us)
