---
title: "USPTO Patents by County"
description: "Patent grants by inventor and assignee location for measuring innovation"
author: "CORI Data Team"
date: "2026-02-02"
categories: ["Dataset", "Innovation"]
tags: ["USPTO", "Patents", "Innovation", "County-Level"]
dataset:
  name: "USPTO PatentsView"
  source: "U.S. Patent and Trademark Office"
  sourceUrl: "https://patentsview.org/"
  accessMethod: "PatentsView API"
  updateFrequency: "Weekly"
  geographicLevel: "County, MSA, State, National"
  dataFormat: ["API", "Bulk Download"]
  featured: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Overview

The United States Patent and Trademark Office (USPTO) grants patents for inventions. PatentsView is a data platform that makes USPTO patent data accessible for research and analysis, including geographic attribution to inventor and assignee locations.

Patent counts serve as an **innovation indicator**, revealing which regions are producing new technologies, products, and processes. Patent data is especially valuable for tracking:

- Regional innovation capacity
- Technology clusters
- University and corporate research output
- Rural innovation ecosystems

::: callout-note
## Why Patent Data Matters

- **Objective innovation measure** - Patents require rigorous examination
- **Long time series** - Data back to 1976
- **Technology detail** - Understand what is being invented
- **Geographic precision** - County-level attribution
- **Linked data** - Citations, inventors, assignees, classifications
:::

## Coverage

::: columns
::: {.column width="50%"}
**Geographic Coverage**

- County
- Metropolitan Statistical Area
- State
- National

(by inventor or assignee location)
:::

::: {.column width="50%"}
**Temporal Coverage**

- 1976-present
- Weekly updates
- New grants published every Tuesday
- Universe: All U.S. utility patents
:::
:::

## Key Variables

### Patent Counts

- **Total Patents** - Count of patents granted
- **Patents by Technology Field** - CPC/IPC classification
- **Patents by Assignee Type** - Corporate, university, government, individual

### Patent Details

- **Patent Number** - Unique identifier
- **Grant Date** - Date patent was granted
- **Filing Date** - Date application was filed (lag: 2-3 years)
- **Inventor Location** - Geographic location of inventors
- **Assignee Location** - Geographic location of patent owner (company, university)

### Citation Metrics

- **Citations Received** - How many later patents cite this patent (measure of influence)
- **Citations Made** - Prior art references

## Access

::: panel-tabset
## PatentsView API

Free API (no authentication required):

```{r}
#| eval: false

# Using patentsview package
library(patentsview)

# Get patents for a specific county in 2023
patents <- search_pv(
  query = qry_funs$and(
    qry_funs$eq(inventor_county = "Washington"),
    qry_funs$eq(inventor_state = "VT"),
    qry_funs$gte(patent_date = "2023-01-01"),
    qry_funs$lte(patent_date = "2023-12-31")
  ),
  fields = c("patent_number", "patent_date", "inventor_city")
)
```

## Bulk Download

Compressed data files:

- [PatentsView Data Download](https://patentsview.org/download/data-download-tables)

## CORI Projects

```{r}
#| eval: false

# From rural-economic-outlook project
# Load script: R/data_import/load_patent.R
# Aggregates to county level, 2014-2024
```
:::

## Data Structure

**County Annual Format:**

- One row per county per year
- GEOID: 5-digit FIPS code
- Patent counts by assignee location

**Typical Variables:**

```
geoid, year, cnt_patent, patents_per_100k_population
```

## Methodology

### Geographic Attribution

Patents can be attributed to locations in multiple ways:

::: columns
::: {.column width="50%"}
**Inventor Location**

Where inventors reside (can be multiple inventors per patent)
:::

::: {.column width="50%"}
**Assignee Location**

Where patent owner is located (company HQ, university)
:::
:::

**CORI uses assignee location** because:

- Reflects where innovation benefits accrue (jobs, revenue)
- More stable over time (inventors may move)
- Aligns with economic development goals

### Fractional Counting

When multiple inventors/assignees exist:

- Each location receives fractional credit
- Example: Patent with 2 inventors in different counties → 0.5 patents counted for each county

### Technology Classification

Patents classified using:

- **CPC (Cooperative Patent Classification)** - 250,000+ classifications
- **IPC (International Patent Classification)** - Global standard
- **NBER Technology Categories** - 37 high-level categories (computers, pharma, etc.)

## Use Cases

### In CORI Projects

| Project | Use Case |
|---------|----------|
| Rural Innovation Analysis | Patent counts in rural vs. metro areas 2014-2024 |
| Technology Clusters | Where is innovation concentrated? |
| University Impact | Patents from rural-serving universities |
| Innovation Capacity | Long-term innovation trends |

### Common Applications

- Economic development strategy
- University tech transfer assessment
- Venture capital investment targeting
- Regional competitiveness analysis
- STEM workforce planning

## Strengths & Limitations

::: columns
::: {.column width="50%"}
### Strengths

- Objective innovation measure
- Long time series (1976+)
- Technology detail
- Geographic precision
- Linked data (citations, etc.)
:::

::: {.column width="50%"}
### Limitations

- Not all innovation patented
- Industry variation
- Quality variation
- Lag from invention (2-3 years)
- Low counts in rural areas
- Assignee location bias
:::
:::

## Important Notes

### Rural Patent Patterns

::: callout-warning
## Low Rural Patent Counts

Rural counties typically have **very low patent counts**:

- 50% of rural counties: 0-1 patents per year
- Patent activity concentrated in:
  - University counties (land-grant institutions)
  - Manufacturing hubs (ag equipment, natural resources)
  - Military installations (defense technology)

**Interpretation:** Patent counts are **not suitable** as a standalone rural innovation measure.
:::

Combine with:

- [Business Dynamics Statistics](/datasets/census-bds/) - Entrepreneurship
- [Business Formation Statistics](/datasets/census-bfs/) - Startup activity
- Industry employment in knowledge-intensive sectors

### Patents vs. Innovation

Patents measure **formal, technical innovation**. Many forms of innovation are not patented:

- Business model innovation
- Process improvements
- Service innovation
- Open-source software

Rural innovation may be **underrepresented** in patent data.

### Citation Analysis

Citation counts reveal patent influence:

::: panel-tabset
## Highly Cited Patents

Foundational innovations with significant impact

## Zero Citations

Limited impact or very recent patents

## Rural Context

Rural patents tend to have **fewer citations** on average, potentially indicating:

- Applied vs. fundamental research
- Narrower market applications
- Lower visibility in patenting community
:::

## Related Datasets

- [Census Business Dynamics Statistics](/datasets/census-bds/) - Business formation (complementary innovation measure)
- [QCEW Employment](/datasets/qcew-employment-wages/) - Employment in R&D-intensive industries
- [American Community Survey](/datasets/american-community-survey/) - STEM occupations

## R Packages

### Recommended Packages

```{r}
#| eval: false

# Install packages
install.packages("patentsview")  # R interface to PatentsView API
```

### Usage Example

```{r}
#| eval: false

library(patentsview)

# Search for patents in a specific technology area
tech_patents <- search_pv(
  query = qry_funs$and(
    qry_funs$gte(patent_date = "2020-01-01"),
    qry_funs$text_phrase(cpc_subgroup_id = "G06N")  # AI/ML patents
  ),
  fields = c("patent_number", "patent_date", "assignee_organization"),
  endpoint = "patents"
)

# Get inventor locations
inventor_locs <- search_pv(
  query = qry_funs$eq(inventor_state = "VT"),
  fields = c("patent_number", "inventor_city", "inventor_county"),
  endpoint = "patents"
)
```

## Additional Resources

- [PatentsView Data User Guide](https://patentsview.org/documentation)
- [Rural Economic Outlook Project](https://github.com/ruralinnovation/data-rural-economic-outlook)

---

**Last Updated:** February 2, 2026
**Data Coverage:** 1976-2024 (CORI uses 2014-2024)
**Contact:** [data@ruralinnovation.us](mailto:data@ruralinnovation.us)
