---
title: "cori.data.bds"
description: "Access and analyze Census Bureau Business Dynamics Statistics (BDS) data covering establishment and job dynamics from 1978-2023"
author: "Drew Rosebush"
date: "2026-01-31"
categories: ["R Package", "Economic Data"]
tags: ["Census", "Business Dynamics", "Employment", "Startups", "Economic Development"]
package:
  name: "cori.data.bds"
  githubUrl: "https://github.com/ruralinnovation/cori.data.bds"
  installCommand: 'devtools::install_github("ruralinnovation/cori.data.bds")'
  status: "stable"
  version: "0.2.0"
  maintainer: "Drew Rosebush"
  featured: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Overview

The `cori.data.bds` package provides tools for accessing and analyzing Census Bureau Business Dynamics Statistics (BDS) data. This comprehensive dataset covers establishment and job dynamics from 1978-2023, including:

- **Establishment dynamics**: Births, deaths, entry rates, exit rates
- **Employment dynamics**: Job creation, destruction, and net employment change
- **Firm dynamics**: Firm counts, deaths, and transitions
- **Multiple breakouts**: By geography, industry (NAICS), firm age, and employment size

::: callout-note
## Why Use This Package

- **Simplified access** - Pre-processed data with human-readable variable names
- **Two data sources** - Fast S3 access or direct Census API pulls
- **Smart filtering** - Reduce data size by 99.98% using aggregation levels
- **Performance optimized** - Efficient data structures for large-scale analysis
:::

## Installation

::: panel-tabset
## GitHub

```{r}
#| eval: false

# Install from GitHub
devtools::install_github("ruralinnovation/cori.data.bds")
```

## Requirements

**Census API Key** (required for pulling data directly):
```{r}
#| eval: false

# Get key at: http://api.census.gov/data/key_signup.html
Sys.setenv(CENSUS_API_KEY = "your_api_key_here")
```

**S3 Access** (optional, for reading/writing processed data):
```{r}
#| eval: false

# Install cori.db package
remotes::install_github("ruralinnovation/cori.db")

# Configure AWS CLI
# aws configure
```
:::

## Quick Start

```{r}
#| eval: false

library(cori.data.bds)

# Read processed data with human-readable names (recommended)
data <- read_bds_from_s3(
  geography = "county",
  vintage = 2023,
  years = 2020:2023,
  aggregation_level = "base",
  data_type = "processed"
)

# Or pull data directly from Census API
raw_data <- pull_bds_raw(years = 2020:2023, geography = "state")

# Get variable definitions
codebook <- get_bds_codebook()
```

## Key Features

### Two Data Access Methods

::: panel-tabset
## S3 (Recommended)

Fast, pre-processed data with human-readable variable names:

```{r}
#| eval: false

read_bds_from_s3(
  geography = "county",
  years = 2020:2023,
  aggregation_level = "base",
  data_type = "processed"
)
```

## Census API

Pull raw data directly from Census Bureau:

```{r}
#| eval: false

pull_bds_raw(years = 2020:2023, geography = "state")
```
:::

### Processed vs. Raw Data

::: columns
::: {.column width="50%"}
**Processed Data** (Recommended):

- Human-readable column names (`establishments` instead of `ESTAB`)
- Industry labels ("Information" instead of "51")
- Age group labels ("0 years" instead of "010")
- Standardized `geoid` column for easy joins
:::

::: {.column width="50%"}
**Raw Data**:

- Original Census variable names
- Requires manual lookup for codes
- Useful for Census API compliance
:::
:::

## Core Functions

### Reading Data

```{r}
#| eval: false

# Read from S3 (fast, efficient)
read_bds_from_s3(
  geography = "county",        # "national", "state", or "county"
  vintage = 2023,              # Latest year in dataset
  years = 2020:2023,           # Optional: filter specific years
  aggregation_level = "base",  # "base", "by_industry", "by_age", "by_size", "all"
  data_type = "processed"      # "processed" or "raw"
)

# Pull directly from Census API
pull_bds_raw(
  years = 2020:2023,
  geography = "state",
  variables = NULL  # Optional: specify variables
)
```

### Writing Data (Admin)

```{r}
#| eval: false

# Write complete dataset to S3 (runs 8-15 hours)
write_bds_to_s3(vintage = 2023, sync_to_s3 = TRUE)

# Create processed data with human-readable names
write_bds_processed_to_s3(vintage = 2023, sync_to_s3 = TRUE)
```

### Utility Functions

```{r}
#| eval: false

# Get codebook
get_bds_codebook()

# Lookup industry names
lookup_naics("51")        # "Information"
lookup_naics("31-33")     # "Manufacturing"

# Lookup age groups
lookup_eage("010")        # "0 years"
lookup_eage("070")        # "6-10 years"

# Lookup employment size categories
lookup_empszes("212")     # "1-4 employees"
lookup_empszes("253")     # "500+ employees"
```

## Aggregation Levels

Filter by aggregation level to reduce data size and improve performance:

| Level | Description | Typical Rows (County, 4 years) | Size |
|-------|-------------|--------------------------------|------|
| `base` | Grand totals only | ~12K | ~1 MB |
| `by_industry` | Industry breakouts | ~240K | ~50 MB |
| `by_age` | Firm age breakouts | ~63K | ~15 MB |
| `by_size` | Employment size breakouts | ~126K | ~30 MB |
| `all` | Everything | ~314K | ~75 MB |

::: callout-important
## Mutual Exclusivity

NAICS (industry), EAGE (age), and EMPSZES (size) are **mutually exclusive** - you can only filter by ONE dimension at a time.
:::

## Common Use Cases

### County Rankings

```{r}
#| eval: false

library(dplyr)

base_2023 <- read_bds_from_s3(
  geography = "county",
  years = 2023,
  aggregation_level = "base",
  data_type = "processed"
)

# Top counties by establishment count
base_2023 |>
  arrange(desc(establishments)) |>
  head(10)
```

### Industry Analysis

```{r}
#| eval: false

industry_data <- read_bds_from_s3(
  geography = "state",
  years = 2020:2023,
  aggregation_level = "by_industry",
  data_type = "processed"
)

# Industry trends
industry_data |>
  filter(year == 2023) |>
  arrange(desc(establishments)) |>
  select(geoid, industry, establishments, employees)
```

### Startup Activity

```{r}
#| eval: false

age_data <- read_bds_from_s3(
  geography = "state",
  years = 2020:2023,
  aggregation_level = "by_age",
  data_type = "processed"
)

# Young firms (0-2 years old)
startups <- age_data |>
  filter(eage_code %in% c('010', '020', '030')) |>
  select(year, geoid, estab_age_group, establishments, firm_deaths_in_12mos)
```

### Small Business Analysis

```{r}
#| eval: false

size_data <- read_bds_from_s3(
  geography = "county",
  years = 2020:2023,
  aggregation_level = "by_size",
  data_type = "processed"
)

# Small businesses (1-19 employees)
small_biz <- size_data |>
  filter(empszes_code %in% c('212', '220', '230'))  # 1-4, 5-9, 10-19
```

## Key Variables

### Processed Data (Human-Readable)

| Variable | Description |
|----------|-------------|
| year | Year (1978-2023) |
| geoid | Geographic identifier (FIPS code) |
| industry | Human-readable industry name |
| estab_age_group | Human-readable age group |
| estab_size_group | Human-readable employment size label |
| establishments | Number of establishments |
| employees | Number of employees |
| estab_entries_in_12mos | Establishments opened |
| estab_exits_in_12mos | Establishments closed |
| estab_entry_rate_in_12mos | Rate of establishment births |
| estab_exit_rate_in_12mos | Rate of establishment exits |
| firms | Number of firms |
| firm_deaths_in_12mos | Firm deaths |
| job_creation | Jobs created from births and expansions |
| job_destruction | Jobs destroyed from closures and contractions |
| net_job_creation | Net employment change |

### Raw Data (Census Variables)

| Variable | Description |
|----------|-------------|
| YEAR | Year (1978-2023) |
| NAICS | Industry code ('00'=all, '51'=Information, etc.) |
| EAGE | Age code ('001'=all, '010'=0 years, etc.) |
| EMPSZES | Employment size code ('001'=all, '212'=1-4 employees, etc.) |
| ESTAB | Number of establishments |
| EMP | Number of employees |
| ESTABS_ENTRY | Establishments opened |
| ESTABS_EXIT | Establishments closed |
| FIRM | Number of firms |
| FIRMDEATH_FIRMS | Firm deaths |
| JOB_CREATION | Jobs created |
| JOB_DESTRUCTION | Jobs destroyed |
| NET_JOB_CREATION | Net jobs |

## Performance Tips

::: callout-tip
## Optimization Strategies

1. **Use processed data** - Pre-computed labels and human-readable names
2. **Use specific aggregation levels** - `base` is 99.98% smaller than `all`
3. **Filter by years** - Reduces data transfer
4. **Start with state or national** - Smaller than county for testing
5. **Test locally first** - Use `sync_to_s3 = FALSE` when writing data
:::

## Data Coverage

- **Time Period**: 1978-2023 (46 years)
- **Geographies**: National, State, County
- **Industries**: All NAICS sectors (2-digit through 4-digit)
- **Firm Ages**: 0 years through 11+ years
- **Employment Sizes**: 1-4 through 1,000+ employees
- **Update Frequency**: Annual (Census releases new data each year)

## Use Cases at CORI

This package is used in:

- Startup ecosystem analysis
- Rural entrepreneurship research
- Economic development studies
- Regional employment trends
- Industry composition analysis

## Related Resources

### Datasets

- [Census Business Dynamics Statistics](/datasets/census-bds/) - Source dataset documentation
- [Census Business Formation Statistics](/datasets/census-bfs/) - Leading indicator
- [QCEW Employment & Wages](/datasets/qcew-employment-wages/) - Quarterly employment data

### External Links

- [Census BDS Homepage](https://www.census.gov/programs-surveys/bds.html)
- [BDS Documentation](https://www.census.gov/programs-surveys/bds/documentation.html)
- [Census API Documentation](https://www.census.gov/data/developers/data-sets/business-dynamics.html)

## Documentation

- [GitHub Repository](https://github.com/ruralinnovation/cori.data.bds)
- [Full README](https://github.com/ruralinnovation/cori.data.bds/blob/main/README.md)
- [Function Documentation](https://github.com/ruralinnovation/cori.data.bds/tree/main/man)
- [Vignettes](https://github.com/ruralinnovation/cori.data.bds/tree/main/vignettes)

---

**Contact:** Drew Rosebush (drew.rosebush@ruralinnovation.us)

For bug reports or feature requests, please open an issue on [GitHub](https://github.com/ruralinnovation/cori.data.bds/issues).
