---
title: "Census Building Permits Survey (BPS)"
description: "Monthly and annual data on housing units authorized by building permits"
author: "CORI Data Team"
date: "2026-02-02"
categories: ["Dataset", "Housing"]
tags: ["Census", "Building Permits", "Housing Supply", "County-Level"]
dataset:
  name: "Building Permits Survey (BPS)"
  source: "U.S. Census Bureau"
  sourceUrl: "https://www.census.gov/construction/bps/"
  accessMethod: "Census API and Bulk Download"
  updateFrequency: "Monthly"
  geographicLevel: "County, Place, MSA, State, National"
  dataFormat: ["CSV", "API", "Excel"]
  featured: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Overview

The Building Permits Survey (BPS) collects data on residential building permits issued by permit-issuing jurisdictions across the United States. Building permits serve as a **leading indicator** of housing supply and construction activity, signaling where new housing development is occurring.

BPS data is widely used to track housing market conditions, assess housing supply constraints, and understand regional growth patterns.

::: callout-note
## Why BPS is Essential

- **High frequency** - Monthly updates for timely tracking
- **Leading indicator** - Precedes construction starts by ~2-3 months
- **Long time series** - Data back to 1980
- **Universal coverage** - All jurisdictions surveyed annually
:::

## Coverage

::: columns
::: {.column width="50%"}
**Geographic Coverage**

- County
- Place
- Metropolitan Statistical Area
- State
- National
:::

::: {.column width="50%"}
**Temporal Coverage**

- 1980-present (monthly)
- Released mid-month for prior month
- Universe: ~20,000 permit-issuing jurisdictions
- Sample: ~1,400 in monthly survey
:::
:::

## Key Variables

### Permit Counts

- **Total Housing Units Authorized** - Number of units approved by building permits
- **Single-Family Units** - Detached single-family homes
- **Multi-Family Units** - Buildings with 2+ units
- **Units in Buildings with 5+ Units** - Larger apartment complexes

### Valuation

- **Construction Value** - Total estimated construction cost (in thousands of dollars)

### Derived Metrics (CORI Projects)

- **Permits per 1,000 Population** - Normalized housing supply indicator
- **Year-over-year Growth Rate** - Housing supply momentum

## Access

::: panel-tabset
## Building Permits Data Portal

Interactive query tool:

- [BPS Data Portal](https://www.census.gov/construction/bps/)

## Bulk Download

Annual files by geography:

- [BPS Datasets](https://www.census.gov/construction/bps/how_the_data_are_collected/index.html)

## Census API

```{r}
#| eval: false

# Using censusapi package
library(censusapi)

permits <- getCensus(
  name = "timeseries/bps/annsa",
  vars = c("UNITS", "POP"),
  region = "county:*",
  time = "2023"
)
```

## CORI Projects

```{r}
#| eval: false

# From rural-economic-outlook project
# Load script: R/data_import/load_bps.R
# Calculates units per 1,000 population
```
:::

## Data Structure

**County Annual Format:**

- One row per county per year
- GEOID: 5-digit FIPS code
- Permit counts and population

**Typical Variables:**

```
geoid, year, units_authorized, population,
units_per_1k_people
```

## Methodology

### Data Collection

BPS operates in two phases:

**1. Monthly Survey (~1,400 jurisdictions):**

- Largest permit-issuing places
- Covers ~80% of all permits
- Released monthly with ~3-week lag

**2. Annual Comprehensive Survey (~20,000 jurisdictions):**

- All known permit-issuing jurisdictions
- Comprehensive coverage
- Used for annual totals and county estimates

### County-Level Estimates

County totals are derived by:

1. Aggregating all permit-issuing jurisdictions within the county
2. Imputing for non-responding jurisdictions
3. Annual reconciliation with comprehensive survey

### Coverage Notes

::: callout-important
## Rural Coverage Considerations

- **Not all counties issue permits** - Some rural areas don't require permits
- **Undercounting in rural areas** - Informal construction, owner-built homes
- **Jurisdictional complexity** - Some places span multiple counties
:::

## Use Cases

### In CORI Projects

| Project | Use Case |
|---------|----------|
| [Rural Economic Outlook](/projects/rural-economic-outlook/) | Housing supply indicator 2007-2024 |
| Cost Pressures Analysis | Tracking housing supply constraints |
| Growth Patterns | Where is housing being built? |
| Rural Housing Supply | Per-capita permit rates in rural vs. metro |

### Common Applications

- Housing market analysis
- Economic development planning
- Infrastructure planning (schools, utilities)
- Demographic projections
- Construction industry tracking

## Strengths & Limitations

::: columns
::: {.column width="50%"}
### Strengths

- High frequency (monthly)
- Leading indicator
- Long time series (1980+)
- Universal coverage
- Administrative data
:::

::: {.column width="50%"}
### Limitations

- Permits ≠ Construction
- Lag to completion (6-18 months)
- Undercounting in rural areas
- No tenure information
- Jurisdictional complexity
- Suppression in small counties
:::
:::

## Important Notes

### Per-Capita Interpretation

Permits per 1,000 population is normalized to account for county size:

```
Units per 1k = (Total units authorized / County population) × 1,000
```

::: panel-tabset
## Typical Values

**High-growth areas:** 5-10+ units per 1,000 population

**Average growth:** 2-4 units per 1,000

**Slow-growth/declining:** 0-2 units per 1,000

## Rural Context

Rural counties typically have **lower permits per capita** than urban/suburban counties, reflecting:

- Slower population growth
- Less demand for new housing
- Informal construction not captured in permits
:::

### Permits as Leading Indicator

Timeline from permit to occupancy:

```
Permit issued → 2-3 months → Construction starts →
6-12 months → Construction completes → Occupancy
```

BPS permits lead [Census Population Estimates](/datasets/census-population-estimates/) by ~12-18 months.

### Relationship to ACS Housing

[ACS housing data](/datasets/american-community-survey/) captures the **stock** of housing (total units, occupancy rates, values). BPS captures the **flow** of new housing (additions to stock).

## Related Datasets

- [Census Population Estimates](/datasets/census-population-estimates/) - Population for per-capita calculations
- [American Community Survey](/datasets/american-community-survey/) - Housing stock, values, occupancy
- [Census Housing Units Estimates](/datasets/census-housing-units/) - Total housing unit counts

## R Packages

### Recommended Packages

```{r}
#| eval: false

# Install packages
install.packages("censusapi")    # Census API wrapper
```

### Usage Example

```{r}
#| eval: false

library(censusapi)

# Get building permits data
permits <- getCensus(
  name = "timeseries/bps/annsa",
  vars = c("UNITS", "VALUE", "POP"),
  region = "county:*",
  time = "2023"
)
```

## Additional Resources

- [BPS Methodology](https://www.census.gov/construction/bps/how_the_data_are_collected/index.html)
- [Rural Economic Outlook Project](https://github.com/ruralinnovation/data-rural-economic-outlook)

---

**Last Updated:** February 2, 2026
**Data Coverage:** 1980-2024 (monthly)
**Contact:** [data@ruralinnovation.us](mailto:data@ruralinnovation.us)
