---
title: "American Community Survey (ACS) 5-Year Estimates"
description: "Comprehensive demographic, social, economic, and housing data from the U.S. Census Bureau"
author: "CORI Data Team"
date: "2026-02-02"
categories: ["Dataset", "Demographics", "Labor Market", "Housing"]
tags: ["Census", "ACS", "Demographics", "County-Level", "Survey"]
dataset:
  name: "American Community Survey (ACS) 5-Year Estimates"
  source: "U.S. Census Bureau"
  sourceUrl: "https://www.census.gov/programs-surveys/acs"
  accessMethod: "Census API and Data Portal"
  updateFrequency: "Annual (5-year rolling averages)"
  geographicLevel: "County, Tract, Block Group, Place"
  dataFormat: ["API", "CSV", "Database"]
  featured: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Overview

The American Community Survey (ACS) is the largest household survey in the United States, collecting detailed demographic, social, economic, and housing information from ~3.5 million households annually. The **5-year estimates** combine data from 5 consecutive years, providing reliable statistics for small geographies like counties and census tracts.

ACS replaced the decennial census "long form" in 2010, providing **annual updates** on topics previously available only every 10 years.

::: callout-note
## Why ACS is Essential

- **Comprehensive coverage** - 40,000+ variables across all topics
- **Annual updates** - More frequent than decennial census
- **Small-area estimates** - County and tract-level data
- **National consistency** - Comparable across all geographies
:::

## Coverage

::: columns
::: {.column width="50%"}
**Geographic Levels**

- County
- Census Tract
- Block Group
- Place
- ZIP Code Tabulation Area
:::

::: {.column width="50%"}
**Temporal Coverage**

- 2005-present (5-year estimates)
- Annual updates
- Universe: All U.S. counties
- Sample: ~3.5M households/year
:::
:::

## Key Variable Categories

The ACS contains 40,000+ variables across 9 major subject areas:

### Demographics
- Age, sex, race, Hispanic origin
- Household composition and relationships
- Fertility and grandparents as caregivers
- Marital history and veteran status

### Education
- Educational attainment (high school, bachelor's, graduate degrees)
- School enrollment
- Field of bachelor's degree

### Labor Force & Employment
- Labor force participation rate
- Employment status
- Unemployment rate
- **Prime-age employment (25-54)** - Key economic indicator
- Industry and occupation (22 detailed categories)

### Income & Poverty
- Median household income
- Per capita income
- Income by race/ethnicity
- Poverty status
- Public assistance receipt
- **Gini index** - Income inequality measure

### Commuting & Work
- **Remote work rate** - Workers working from home
- Means of transportation to work
- Travel time to work
- Place of work

### Housing
- Homeownership rate
- Median home value
- Median rent
- **Housing cost burden** - % paying >30% of income for housing
- Vacancy rates
- Year structure built

### Internet Access
- Broadband subscription rates
- Computer ownership
- Internet access by type

### Health Insurance
- Insurance coverage rate
- Coverage type (employer, Medicaid, Medicare, etc.)

### Migration
- Residence 1 year ago
- Geographic mobility

## CORI Standard Variables (88 Variables)

From the rural-economic-outlook project, CORI regularly uses 88 ACS variables:

::: panel-tabset
## Labor Market

Core metrics for workforce analysis:

- Prime-age employment rate (25-54 years)
- Self-employment rate
- Remote work rate (working from home)
- Unemployment rate
- Labor force participation rate

## Occupations

22 detailed occupation categories:

- Management
- Business and financial operations
- Computer and mathematical
- Architecture and engineering
- Life, physical, and social science
- Community and social service
- Legal
- Education, training, and library
- Arts, design, entertainment, sports, media
- Healthcare practitioners and technical
- Healthcare support
- Protective service
- Food preparation and serving
- Building and grounds cleaning and maintenance
- Personal care and service
- Sales and related
- Office and administrative support
- Farming, fishing, and forestry
- Construction and extraction
- Installation, maintenance, and repair
- Production
- Transportation and material moving

## Economic Measures

- Median household income
- Per capita income
- Gini index (income inequality)
- Poverty rate

## Housing

- Homeownership rate
- Median home value
- Median gross rent
- Owner housing cost burden (>30% income)
- Renter housing cost burden (>30% income)

## Education

- Bachelor's degree or higher attainment (%)
- High school diploma or higher (%)
:::

## Access

### Census API

Free API access with registration at [Census API Key Signup](https://api.census.gov/data/key_signup.html):

::: panel-tabset
## tidycensus Package

```{r}
#| eval: false

# Using tidycensus package
library(tidycensus)

# Get median household income for all counties in 2023
income_data <- get_acs(
  geography = "county",
  variables = "B19013_001",  # Median household income
  year = 2023,
  survey = "acs5"
)

# Get multiple variables with descriptive names
variables <- c(
  median_income = "B19013_001",
  employment_rate = "B23025_004",
  remote_work = "B08006_017"
)

acs_data <- get_acs(
  geography = "county",
  variables = variables,
  year = 2023
)
```

## Data Portal

Interactive download tool:

- [data.census.gov](https://data.census.gov)

## CORI Projects

```{r}
#| eval: false

# From rural-economic-outlook project
# Load script: R/data_import/load_acs5_slim.R
# Pulls 88 standardized variables for 2013, 2018, 2023
```
:::

## Data Structure

**Standard Format:**

- One row per geography per variable
- GEOID: 5-digit (county), 11-digit (tract), 12-digit (block group)
- Estimates with margins of error (MOE)

**Typical Variables:**

```
geoid, year, variable, estimate, moe,
variable_label, concept
```

## Methodology

### Rolling 5-Year Averages

The 5-year ACS (e.g., 2019-2023) represents data collected over 60 months:

- Provides **larger sample size** for reliable small-area estimates
- Data represents **average conditions** over the period, not a specific year
- Trade-off: Less timely than 1-year estimates but more precise

### Survey Design

- **Continuous monthly sampling** throughout the year
- ~3.5 million addresses sampled annually
- Internet, mail, phone, and in-person interviews
- Household-based (not institutional population)

### Margins of Error (MOE)

All ACS estimates include MOE at 90% confidence level:

::: callout-important
## Understanding Margins of Error

- Small geographies have larger MOE
- Use MOE to determine statistical significance
- Aggregate across geographies to reduce MOE

**Example:** If estimate = 1,000 and MOE = 200, we are 90% confident the true value is between 800 and 1,200.
:::

## Use Cases

### In CORI Projects

| Project | Use Case |
|---------|----------|
| [Rural Economic Outlook](/projects/rural-economic-outlook/) | Labor market, housing cost burden, remote work trends |
| Occupation Analysis | Tech jobs, professional occupations in rural areas |
| Housing Affordability | Cost burden by rural/metro classification |
| Educational Attainment | Bachelor's degree rates and workforce quality |
| Income Inequality | Gini index trends over time |

### Common Applications

- Community needs assessment
- Grant applications (affordable housing, workforce development)
- Economic development planning
- Market analysis
- Transportation planning (commuting patterns)

## Strengths & Limitations

::: columns
::: {.column width="50%"}
### Strengths

- Comprehensive coverage (40,000+ variables)
- Small-area estimates (county and tract level)
- Annual updates
- Detailed demographics
- Occupation detail (500+ codes)
- National consistency
:::

::: {.column width="50%"}
### Limitations

- Survey-based (sampling error/MOE)
- 5-year lag (2019-2023 released Dec 2024)
- Rolling averages (not single-year snapshot)
- Small numbers unreliable (high MOE)
- Top-coding (high incomes/values capped)
- Disclosure avoidance (noise added)
:::
:::

## Important Notes

### Variable Consistency Across Years

::: callout-tip
## Documented Consistency for 2013, 2018, 2023

CORI has verified variable consistency:

- **Underlying variable codes (B-series) remain stable** across years
- **Label formatting changed:** "!!" separators (2013) became "::" (2018+)
- **Terminology changes:** "at home" → "from home" (work location)
- **Data is comparable** despite cosmetic label changes

See documentation:

- `/data/acs_variable_check.csv` - Verification of 88 variables
- `/data/acs_variable_problems.csv` - Documented label inconsistencies (no data impact)
:::

### Broadband Data

**Note:** CORI uses [FCC National Broadband Map](/datasets/fcc-broadband/) instead of ACS broadband variables because:

- FCC provides location-level precision (addresses)
- ACS broadband is household survey (less precise)
- FCC updates more frequently

### Cost Burden Handling

Housing cost burden variables:

- Calculate as % of households paying >30% of income
- Exclude households with no/negative income (undefined burden)
- Owner and renter burdens calculated separately

## Related Datasets

- [Census Population Estimates](/datasets/census-population-estimates/) - Annual population (no survey error)
- [QCEW Employment & Wages](/datasets/qcew-employment-wages/) - Administrative employment data
- [FCC National Broadband Map](/datasets/fcc-broadband/) - Preferred broadband source

## R Packages

### Recommended Packages

```{r}
#| eval: false

# Install packages
install.packages("tidycensus")   # User-friendly API wrapper
install.packages("censusapi")    # General Census API access
```

### Usage Example

```{r}
#| eval: false

library(tidycensus)

# Set API key (get from census.gov)
census_api_key("YOUR_KEY_HERE", install = TRUE)

# Get ACS data
acs_data <- get_acs(
  geography = "county",
  variables = c(
    median_income = "B19013_001",
    poverty_rate = "B17001_002"
  ),
  year = 2023,
  survey = "acs5",
  state = NULL  # All states
)
```

## Additional Resources

- [ACS Technical Documentation](https://www.census.gov/programs-surveys/acs/technical-documentation.html)
- [ACS Handbook for Data Users](https://www.census.gov/programs-surveys/acs/library/handbooks/general.html)
- [Rural Economic Outlook Project](https://github.com/ruralinnovation/data-rural-economic-outlook)

---

**Last Updated:** February 2, 2026
**Data Coverage:** 2005-2023 (5-year estimates)
**Contact:** [data@ruralinnovation.us](mailto:data@ruralinnovation.us)
