---
title: "USDA ERS County Typology Codes"
description: "Economic and demographic classifications for understanding county characteristics"
author: "CORI Data Team"
date: "2026-02-02"
categories: ["Dataset", "Rural Classification"]
tags: ["USDA", "Economic Specialization", "Persistent Poverty", "County-Level"]
dataset:
  name: "USDA Economic Research Service County Typology Codes"
  source: "U.S. Department of Agriculture, Economic Research Service"
  sourceUrl: "https://www.ers.usda.gov/data-products/county-typology-codes/"
  accessMethod: "Direct Download"
  updateFrequency: "Periodic (last updated: 2025 edition)"
  geographicLevel: "County"
  dataFormat: ["Excel", "CSV"]
  featured: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Overview

The USDA Economic Research Service (ERS) County Typology Codes classify counties into economic and demographic categories based on various characteristics. These classifications help researchers and policymakers:

- Understand county economic specialization
- Identify persistent challenges (poverty, education, employment)
- Stratify analysis by county type
- Target policies to specific county contexts

The typology provides a **static snapshot** of county characteristics, typically updated every 10 years following the decennial census.

::: callout-note
## Why Typology Codes Matter

- **Mutually exclusive categories** - Every county has one economic type
- **Research-validated** - Widely used in rural studies
- **Persistent poverty flag** - Identifies long-term challenges
- **Multiple dimensions** - Economic type + social/demographic flags
- **Free and accessible** - Easy download, no API needed
:::

## Coverage

::: columns
::: {.column width="50%"}
**Geographic Coverage**

- All 3,100+ U.S. counties
:::

::: {.column width="50%"}
**Temporal Coverage**

- 2025 edition (based on 2015-2023 data)
- Previous editions: 2015, 2004, 1989
- Updated roughly decennial
:::
:::

## Key Classifications

### Economic Typology (6 Mutually Exclusive Types)

Counties classified by dominant economic activity:

::: panel-tabset
## Farming-Dependent

Farming contributed ≥25% of county earnings or ≥16% of employment (averaged 2017-2019)

## Mining-Dependent

Mining contributed ≥13% of earnings or ≥8% of employment

## Manufacturing-Dependent

Manufacturing contributed ≥25% of earnings or ≥18% of employment

## Government-Dependent

Federal/State Government contributed ≥14% of earnings

## Recreation

Recreation employment ≥11% of total employment OR per capita personal income from recreation ≥11% of county income

## Nonspecialized

Does not meet criteria for any specialized category (residual category)

**Note:** ~60-70% of counties are nonspecialized
:::

### Persistent Poverty (Binary Flag)

Counties where **≥20% of population was poor** in each of the last 4 decennial censuses:

- 1980, 1990, 2000, 2010 (for 2015 edition)
- 1990, 2000, 2010, 2020 (for 2025 edition)

Identifies counties with **long-term, structural poverty**, not temporary economic shocks.

### Persistent Child Poverty (Binary Flag)

Counties where **≥20% of children (age 0-17) were poor** in each of the last 4 censuses.

Child poverty often more severe and has longer-lasting impacts than adult poverty.

### Low Education (Binary Flag)

Counties where:

- **≥20% of adults (25-64) lack high school diploma** AND
- **<20% have bachelor's degree or higher**

Measured as of 2015-2019 ACS (2025 edition).

### Low Employment (Binary Flag)

Counties with **employment/population ratio <55%** for working-age adults (21-64), averaged 2017-2019.

Indicates weak labor market attachment beyond unemployment (includes discouraged workers, disability).

### Population Loss (Binary Flag)

Counties that **lost population** between 2010-2020 censuses.

Identifies long-term decline, not just short-term fluctuation.

## Access

::: panel-tabset
## Direct Download

Excel and CSV files:

- [USDA ERS County Typology](https://www.ers.usda.gov/data-products/county-typology-codes/)

## CORI Projects

```{r}
#| eval: false

# From rural-economic-outlook project
# Load script: R/data_import/load_ers_typology.R
# 2025 edition, direct CSV download
```
:::

## Data Structure

**County-Level (One Row per County):**

```
geoid, county_name, state_abbr,
typology_2025, farming, mining, manufacturing,
government, recreation, nonspecialized,
persistent_poverty, persistent_child_poverty,
low_education, low_employment, population_loss
```

**Binary flags:** 1 = meets criteria, 0 = does not meet criteria

## Methodology

### Economic Typology Classification

Counties assigned to types using hierarchical rules:

1. Check farming criteria → if met, assign "Farming"
2. If not farming, check mining → if met, assign "Mining"
3. Continue through manufacturing, government, recreation
4. If none met, assign "Nonspecialized"

**Data Sources:**

- BEA Regional Economic Accounts (earnings by industry)
- BLS QCEW (employment by industry)
- Census County Business Patterns (employment)

### Persistent Poverty Methodology

Using decennial census and ACS data:

- Poverty rate calculated for all persons in county
- Must meet 20% threshold in **all 4** censuses to qualify
- Captures structural poverty, not recession-related spikes

::: callout-important
## Geographic Distribution

~15% of U.S. counties meet persistent poverty criteria. Concentrated in:

- **Black Belt** (Southeast)
- **Appalachia**
- **Native American reservations**
- **Texas border counties**
:::

### Thresholds and Sensitivities

Thresholds (e.g., 25% for farming, 20% for persistent poverty) are **somewhat arbitrary**:

- Counties just below thresholds excluded
- Use as rough classification, not precise boundaries

## Use Cases

### In CORI Projects

| Project | Use Case |
|---------|----------|
| [Rural Economic Outlook](/projects/rural-economic-outlook/) | Stratify analysis by economic type |
| Manufacturing Belt Analysis | Identify manufacturing-dependent rural counties |
| Persistent Poverty | Target analysis to structurally disadvantaged counties |
| Economic Resilience | Do recreation counties weather recessions differently? |

### Common Applications

- Targeting federal programs (USDA, ARC, EDA)
- Economic development strategy (diversification for specialized economies)
- Workforce development (low education counties)
- Research stratification (control for county type)

## Strengths & Limitations

::: columns
::: {.column width="50%"}
### Strengths

- Mutually exclusive categories
- Research-validated
- Persistent poverty flag
- Multiple dimensions
- Free and accessible
:::

::: {.column width="50%"}
### Limitations

- Static classification
- Arbitrary thresholds
- Mutually exclusive types (no mixed)
- No temporal variation
- Lagged data (2025 uses 2019 ACS)
:::
:::

## Important Notes

### Economic Typology vs. Rural Definition

::: callout-warning
## Different Concepts

ERS County Typology ≠ rural definition:

- **Rural definition** (CBSA, RUCA) = urban vs. rural
- **Economic typology** = type of economy

**Many metro counties are farming/manufacturing-dependent.** Use [ruraldefinitions package](/packages/ruraldefinitions/) for rural classification.
:::

### Nonspecialized Counties (Majority)

~60-70% of counties are "Nonspecialized":

- Diversified economies (no single sector dominant)
- OR economies too small/weak to meet any threshold

**Nonspecialized ≠ unimportant.** Many thriving counties have diversified economies.

### Persistent Poverty Concentration

Persistent poverty counties are **geographically concentrated**. These regions face **structural challenges** beyond economic cycles:

- Historical discrimination and disinvestment
- Limited infrastructure
- Lower educational attainment
- Geographic isolation

### Recreation Counties

Recreation typology captures:

- Tourism employment (hotels, restaurants, entertainment)
- Outdoor recreation (skiing, beaches, national parks)
- Second-home income

**Rural recreation counties** often have:

- Higher property values
- Seasonal employment
- Income inequality (wealthy visitors vs. low-wage service workers)

### Using Typology for Analysis

::: callout-tip
## Best Practices

1. **Stratify models** - Run separate regressions for farming vs. manufacturing counties
2. **Control variables** - Include typology dummies in regressions
3. **Descriptive analysis** - Compare outcomes across types (e.g., "Do farming counties have lower broadband adoption?")
4. **Geographic mapping** - Visualize spatial patterns of economic specialization
:::

## Related Datasets

- [ruraldefinitions Package](/packages/ruraldefinitions/) - Rural/urban classification (complementary to economic type)
- [QCEW Employment](/datasets/qcew-employment-wages/) - Detailed industry employment for custom classifications
- [BEA Regional Data](/datasets/bea-real-gdp/) - Industry earnings data underlying typology

## R Packages

### Custom Scripts

```{r}
#| eval: false

# No dedicated R package
# CORI uses custom scripts for loading

# From rural-economic-outlook project
source("R/data_import/load_ers_typology.R")

# For rural classification, use:
# install.packages("ruraldefinitions")
library(ruraldefinitions)
```

## Additional Resources

- [ERS Typology Documentation](https://www.ers.usda.gov/data-products/county-typology-codes/descriptions-and-maps/)
- [Rural Economic Outlook Project](https://github.com/ruralinnovation/data-rural-economic-outlook)

---

**Last Updated:** February 2, 2026
**Data Coverage:** 2025 edition (based on 2015-2023 data)
**Contact:** [data@ruralinnovation.us](mailto:data@ruralinnovation.us)
